Emergence of the Southeast Asian islands as a driver for Neogene cooling

### 2020-07-31 Manuscript/Manuscript.tex
update manuscript with the latest

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Yuem Park}
\author[a]{Pierre Maffre} 
\author[b]{Yves Godd\'eris}
\author[c]{Francis A. Macdonald}
\author[c]{Eliel A. Anttila}
\author[a]{Nicholas L. Swanson-Hysell}

@@

% Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the significance statement is to explain the relevance of the work in broad context to a broad readership. The significance statement appears in the paper itself and is required for all research papers.
\significancestatement{The Southeast Asian islands are a hotspot of \COtwo consumption via silicate weathering in the present day. Since $\sim$15 million years ago, these islands have been increasing in size at the same time that Earth's climate has been cooling. Here we test the hypothesis that global cooling over the past 15 million years has been driven by tectonic emergence of the Southeast Asian islands. To do so, we improve a silicate weathering model, then couple it to a climate model. We find that the majority of cooling over the past 15 million years can be explained by emergence of the Southeast Asian islands.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Author contributions: Y.P., P.M., Y.G., and N.L.S.-H. designed and implemented the GEOCLIM model. F.A.M. and E.A.A. constructed the paleoshoreline database. Y.P. and P.M. executed the model and analyzed the data in consultation with N.L.S.-H and Y.G.. Y.P., P.M., Y.G., F.A.M., and N.L.S.-H. wrote the manuscript.}
\authordeclaration{The authors declare no competing interest.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: yuempark@berkeley.edu}

@@

% Please provide an abstract of no more than 250 words in a single paragraph. Abstracts should explain to the general reader the major contributions of the article. References in the abstract must be cited in full within the abstract itself and cited in the text.
\begin{abstract}
Steep topography, a tropical climate, and mafic lithologies contribute to efficient chemical weathering and carbon sequestration in the Southeast Asian islands. Ongoing arc-continent collision between the Sunda-Banda arc system and Australia has increased the area of subaerially-exposed land in the region since the mid-Miocene. Concurrently, Earth's climate has cooled since the Miocene Climatic Optimum, leading to growth of the Antarctic ice sheet and onset of Northern Hemisphere glaciation. We seek to evaluate the hypothesis that the emergence of the Southeast Asian islands played a significant role in driving this cooling trend through increasing global weatherability. To do so, we have compiled paleoshoreline data and incorporated them into GEOCLIM, which couples a global climate model to a silicate weathering model with spatially resolved lithology. We find that without the increase in area of the Southeast Asian islands over the Neogene, atmospheric \pCOtwo would have been significantly higher than present day values, remaining above the levels necessary for initiating Northern Hemisphere ice sheets.
\end{abstract}

@@

\dropcap{T}he Southeast Asian islands (SEAI) have an out-sized contribution to modern chemical weathering fluxes. The confluence of steep topography, a warm and wet tropical climate, and the presence of mafic lithologies results in high fluxes of Ca and Mg cations and associated \COtwo consumption \cite{Gaillardet1999a, Hartmann2009a, Milliman2013a, Hartmann2014a}. There has been a significant increase in the area of subaerially-exposed land within the region since the mid-Miocene associated with ongoing arc-continent collision between Australia and the Sunda-Banda arc system \cite{Molnar2015a, Hall2017a, Macdonald2019a}. Concurrently, after the Miocene Climatic Optimum, a cooling trend began ca. 15~Ma and accelerated over the past 4 million years (m.y.) leading to the development of Northern Hemisphere ice sheets \cite{Shackleton1984a, Zachos2001a}. Many hypotheses have been proposed to explain this cooling trend including changes in ocean/atmosphere circulation \cite{Haug1998a, Shevenell2004a, Molnar2015a}, a decrease in volcanic degassing \cite{Berner1983a}, or uplift in the Himalaya \cite{Raymo1988a, Galy2007a}. Here we seek to evaluate the hypothesis that SEAI emergence was a significant factor in driving long-term climatic cooling over the Neogene.

Over geologic time-scales, \COtwo enters Earth's ocean--atmosphere system primarily via volcanism and metamorphic degassing, and leaves primarily through the chemical weathering of silicate rocks, with a smaller sink through organic carbon burial \cite{Kump2000a}. Chemical weathering delivers alkalinity and cations to the ocean which drives carbon sequestration through carbonate precipitation. Steady-state \pCOtwo is set at the \pCOtwo level at which \COtwo sinks are equal to sources. As \COtwo sinks are removed and \pCOtwo rises, temperature increases and the hydrological cycle is invigorated, causing other weathering sinks to increase until a new steady-state is achieved at higher \pCOtwo \cite{Kump1997a}. Conversely, as regions with high carbon sequestration potential emerge on Earth's surface, \pCOtwo decreases resulting in other weathering sinks to diminish leading to a new lower steady-state \pCOtwo.

Topography, climate, and lithology all effect chemical weathering. High-relief regions lack extensive regolith development, and thus tend to have reaction-limited weathering regimes that are more prone to adjust when climate changes \cite{Gabet2009a, West2012a, Maher2014a}. High physical erosion rates contribute to high chemical weathering fluxes in these high-relief regions \citep{Godderis2017b}. In warm and wet regions, mineral dissolution kinetics are faster leading to enhanced chemical weathering \cite{Lasaga1994a, West2012a}. Mafic rocks have higher Ca and Mg concentrations and dissolution rates than felsic rocks, and thus have the potential to more efficiently sequester carbon through silicate weathering \cite{Dessert2003a}. These factors have lead to the proposal that arc-continent collisions, which create steep landscapes of mafic lithologies, within the tropical rain belt have been important in enhancing global weatherability, lowering atmospheric \pCOtwo, and initiating glacial climate over the past 520~m.y. \cite{Jagoutz2016a, Swanson-Hysell2017a, Macdonald2019a} and perhaps in the Neoproterozoic as well \cite{Park2019b}.

Quantitatively estimating the magnitude of decrease in steady-state \pCOtwo associated with the emergence of a region with a high carbon sequestration potential, such as the SEAI, requires constraints on changing tectonic context and accounting of associated feedbacks. As this region emerges, the total global silicate weathering flux will briefly exceed the volcanic degassing flux, causing \pCOtwo to initially decline until a new steady-state is established where the total magnitude of the \COtwo sinks is the same as before the change. However, the sensitivity of the silicate weathering flux in any particular location to this change in \pCOtwo is variable and dependent on the specific topography, climate, and lithology at that location. Furthermore, how regional climate responds to this change in \pCOtwo is itself spatially-variable. Therefore, the magnitude of \pCOtwo change that is required to balance the total global silicate weathering flux with the volcanic degassing flux will depend on the specific spatial distribution of topography, climate, and lithology at the time of emergence. As a result, any attempt to meaningfully estimate the decrease in steady-state \pCOtwo associated with emergence of the SEAI must model spatially-resolved climatology and silicate weathering fluxes in tandem and account for the spatial distribution of the factors that affect these inter-connected systems.

@@

To estimate the decrease in steady-state \pCOtwo associated with the increase of subaerially-exposed land area in the SEAI, we use the global spatially-resolved GEOCLIM model \cite{Godderis2017c}. GEOCLIM estimates changes in steady-state \pCOtwo associated with coupled changes in erosion, chemical weathering, and climatology by linking a silicate weathering model to climate model runs at multiple \pCOtwo levels.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{Figures/regolith_schematic.pdf}
    \caption{A schematic representation of the silicate weathering component of GEOCLIM in a single profile at steady-state. A rock particle leaves the unweathered bedrock with production rate $P_{r}$, and transits vertically through a regolith of height $h$. Regolith production and physical erosion ($E_{p}$) are equal at steady-state. As a particle transits upwards, some fraction of the primary phases ($x$) are chemically weathered ($W$), with the flux of Ca+Mg being $W$ multiplied by the concentration of Ca+Mg in unweathered bedrock ($\chi_{CaMg}$). Details of the formulation for the silicate weathering component of GEOCLIM can be found in \MM.}
    \label{fig:regolith_schematic}
\end{figure}

The silicate weathering component of GEOCLIM calculates \COtwo consumption resulting from silicate weathering for subaerially-exposed land. We assume that Ca and Mg are the only cations that consume \COtwo over geologic time-scales, such that each mole of Ca or Mg that is dissolved by silicate weathering consumes one mole of \COtwo. In previous versions of the model, silicate weathering was a function of temperature and runoff only, and all bedrock was assigned identical chemical compositions \cite{Godderis2017c}. More recent versions of GEOCLIM implement regolith development and soil shielding (Fig. \ref{fig:regolith_schematic}), which introduces a dependence on erosion rate (and therefore topographic slope) \cite{Maffre2018a}. While this introduction of regolith development into GEOCLIM is important for assessing the impact of tropical arc-continent collisions on \pCOtwo, the relatively high Ca+Mg concentration in arc rocks relative to other lithologies must also be considered.

@@

The values of four parameters within the silicate weathering component that modify the dependence of silicate weathering on temperature, runoff, and regolith thickness are poorly constrained. Rather than prescribing values, we allow these parameters and the Ca+Mg concentration of metamorphic and siliciclastic lithologies to vary within reasonable bounds (\SI; Table S2). This approach results in 93,600 unique combinations of these six parameters. For each combination, we compute spatially-resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available \cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating parameter combinations that result in low $r^{2}$, 573 parameter combinations remain (\SI; Fig. S2). The resulting global \COtwo consumption of these model runs all overlap with independently-derived estimates of the global \COtwo degassing flux \cite{Gerlach2011a} as they should given that the long-term carbon cycle is in steady state (\SI; Fig. S2).

@@

Having calibrated the silicate weathering component of GEOCLIM, we use it to estimate the decrease in steady-state \pCOtwo associated with SEAI emergence. For the climate model component, we use temperature and runoff from a subset of the GFDL CM2.0 experiments \cite{Delworth2006b} (\SI). These experiments are well-suited for this analysis because all non-\COtwo forcings are held constant at values representative of pre-industrial conditions, allowing the effect of changing \pCOtwo on climatology to be isolated. Furthermore, the experiments were run long enough for the final system to approximate steady-state.

@@

\begin{figure*}[h]
    \centering
    \includegraphics[width=1\linewidth]{Figures/scenario_pCO2.pdf}
    \caption{Steady-state \pCOtwo estimates from GEOCLIM for the various scenarios discussed in the text. For each of the seven scenarios, each point represents an estimate from one of the 573 unique parameter combinations that resulted in reasonable total global \COtwo consumption and most closely matched estimates of present-day \COtwo consumption in 80 watersheds around the world (\SI). The box encloses the middle 50\% of the \pCOtwo estimates (i.e. the interquartile range), and the notch represents the median with its 95\% confidence interval. The whiskers extend to the 2.5 and 97.5 percentile values. Glaciation thresholds \cite{DeConto2008a} are shown on the x-axis.}
    \label{fig:scenario_pCO2}
\end{figure*}

To determine the position of SEAI paleoshorelines over the past 15~m.y., we use terrestrial and marine sedimentary deposits (Fig. \ref{fig:shoreline_growth}; \SI). The paleoshoreline data indicate that the Sunda-Banda Arc and New Guinea are primarily responsible for the increase in area since 15~Ma. Exhumation of the modern Sunda-Banda Arc is the result of ongoing arc-continent collision with the Australian Plate \cite{Harris2006a}. Of the major islands in this arc, only Sumatra, Belitung, Bangka, Java, Bali, and Flores were subaerially exposed before 5~Ma. However, most of Sumatra and Java along with the non-volcanic islands of the Outer Banda Arc were elevated above sea level after 5~Ma \cite{Hall2013b}. In New Guinea, emergence in the mid-Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{Cloos2005a}, which drove exhumation of the Irian-Marum-April Ophiolite Belt. Exhumation accelerated over the past 4~m.y. in the New Guinea Central Range due to slab-breakoff and buoyant uplift, and in eastern New Guinea due to jamming of the north-dipping subduction zone \cite{Cloos2005a}. These tectonic drivers and others throughout the region lead to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C). We also include changes in areas of presently-submerged continental shelves such as the Sunda Shelf that were previously exposed (\SI; Fig. S5).

We use GEOCLIM to estimate \pCOtwo associated with the reconstructed subaerial extent of the SEAI at ca. 15, 10, and 5~Ma (``paleo-SEAI'' scenarios; Fig. \ref{fig:scenario_pCO2}). Because we use a climate model forced with modern geography, the position of the tectonic blocks remain fixed. Although there has been motion of these tectonic blocks since 15~Ma, they have remained within tropical latitudes such that this fixed scenario is a good approximation of the paleogeography (\SI; Fig. S8). We also test an end-member scenario, in which all islands associated with arc-continent collision in the region are removed (``removed SEAI'' scenario).

@@

Using the 573 unique parameter combinations, the ``paleo-SEAI'' scenarios resulted in 526--678~ppm for 15~Ma, 457--516~ppm for 10~Ma, and 391--434~ppm \pCOtwo for 5~Ma (Fig. \ref{fig:scenario_pCO2}). These results indicate a progressive decrease in \pCOtwo over the Neogene associated with the emergence of the SEAI, and suggest that without this emergence, pre-industrial \pCOtwo would have been $\sim$526--678~ppm.

Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation \cite{Foster2017a}. The \pCOtwo values from the 5~Ma paleo-SEAI scenario overlap with many proxy-based estimates as well as values that emerge from hybrid climate model data assimilation approaches \cite{van-de-Wal2011a, Berends2020a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene (\SI; Fig. S9), indicating that the increase in subaerially-exposed land area and tectonic topography of the SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo associated with SEAI emergence are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

The results of our SEAI scenarios highlight the importance of the combination of topography, runoff, and lithology in setting Earth's climate state. To independently explore the effect of the modern-day surface exposure of low-relief basaltic lavas on steady-state \pCOtwo \cite{Kent2013a}, we replace mafic volcanics associated with the Deccan Traps, Ethiopian Traps, and Columbia River Basalts with the Ca+Mg concentration of bulk continental crust in GEOCLIM (Fig. \ref{fig:scenario_pCO2}). The resulting \pCOtwo is $<$500~ppm, indicating that, while the presence of mafic rocks in these igneous provinces modulates steady-state \pCOtwo as has been suggested to be important for Paleogene cooling \cite{Kent2013a}, its contribution is less than that of the SEAI, which have high-relief and a wet tropical climate in conjunction with high Ca+Mg lithologies.

Previous work has estimated that the decrease in \pCOtwo since 5~Ma associated with the emergence of the SEAI and enhanced silicate weathering is $\sim$19~ppm \cite{Molnar2015a}, in which case SEAI emergence would be a relatively minor contributor to Neogene cooling. This 19~ppm estimate was obtained using a 0-D equation that assumes a direct linear relationship between mean global temperature and changes in weathering-rate-weighted land area, scaled by a factor that is intended to account for the influence of both runoff and temperature. \pCOtwo was then estimated from the calculated temperature using a simple energy balance equation. However, the relationship between mean global temperature (or \pCOtwo) and weathering-rate-weighted land area is not linear. Furthermore, this simple linear relationship ignores spatial variability in topography and climatology, and only crudely accounts for spatial variability in lithology. In fact, the 19~ppm estimate is closer in magnitude to the decrease in \pCOtwo that we estimate if mafic volcanics associated with the Deccan Traps (a relatively flat area outside of the warm and wet tropics) are replaced with the Ca+Mg concentration of bulk continental crust (22--70 ppm; Fig. \ref{fig:scenario_pCO2}). The significant difference in steady-state \pCOtwo estimated between the ``removed Deccan Traps'' scenario and the ``paleo-SEAI'' scenarios (Fig. \ref{fig:scenario_pCO2}) demonstrates that considering changes in the spatial distribution of lithologies alone is not adequate for estimating changes in steady-state \pCOtwo. Instead, spatially-varying topography and climatology significantly modulates silicate weathering rates, and must be accounted for when estimating \pCOtwo change associated with paleogeographic change.

@@

\section*{The Global Carbon Cycle}

>>>>

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Yuem Park}
\author[a]{Pierre Maffre}
\author[b]{Yves Godd\'eris}
\author[c]{Francis A. Macdonald}
\author[c]{Eliel S. C. Anttila}
\author[a]{Nicholas L. Swanson-Hysell}

@@

% Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the significance statement is to explain the relevance of the work in broad context to a broad readership. The significance statement appears in the paper itself and is required for all research papers.
\significancestatement{The Southeast Asian islands are a modern-day hotspot of \COtwo consumption via silicate weathering. Since $\sim$15 million years ago, these islands have been increasing in size at the same time that Earth's climate has been cooling. Here we test the hypothesis this global cooling could have been driven by tectonic emergence of the Southeast Asian islands. Using a new compilation of paleoshorelines in conjunction with a coupled silicate weathering and climate model, we find that this emergence is associated with a large decrease in \pCOtwo. Without these changes in tropical island paleogeography, there would not have been large Northern Hemisphere ice sheets as a defining feature of Earth's climate over the past 3 million years.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Author contributions: Y.P., P.M., Y.G., and N.L.S.-H. designed and implemented the GEOCLIM model. F.A.M. and E.S.C.A. constructed the paleoshoreline database. Y.P. and P.M. executed the model and analyzed the data in consultation with N.L.S.-H and Y.G.. Y.P., P.M., Y.G., F.A.M., and N.L.S.-H. wrote the manuscript.}
\authordeclaration{The authors declare no competing interest.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: yuempark@berkeley.edu}

@@

% Please provide an abstract of no more than 250 words in a single paragraph. Abstracts should explain to the general reader the major contributions of the article. References in the abstract must be cited in full within the abstract itself and cited in the text.
\begin{abstract}
Steep topography, a tropical climate, and mafic lithologies contribute to efficient chemical weathering and carbon sequestration in the Southeast Asian islands. Ongoing arc-continent collision between the Sunda-Banda arc system and Australia has increased the area of subaerially exposed land in the region since the mid-Miocene. Concurrently, Earth's climate has cooled since the Miocene Climatic Optimum, leading to growth of the Antarctic ice sheet and onset of Northern Hemisphere glaciation. We seek to evaluate the hypothesis that the emergence of the Southeast Asian islands played a significant role in driving this cooling trend through increasing global weatherability. To do so, we have compiled paleoshoreline data and incorporated them into GEOCLIM, which couples a global climate model to a silicate weathering model with spatially resolved lithology. We find that without the increase in area of the Southeast Asian islands over the Neogene, atmospheric \pCOtwo would have been significantly higher than pre-industrial values, remaining above the levels necessary for initiating Northern Hemisphere ice sheets.
\end{abstract}

@@

\dropcap{T}he Southeast Asian islands (SEAI) have an out-sized contribution to modern chemical weathering fluxes relative to its area. The confluence of steep topography, a warm and wet tropical climate, and the presence of mafic lithologies results in high fluxes of Ca and Mg cations in the dissolved load and associated \COtwo consumption \cite{Gaillardet1999a, Hartmann2009a, Milliman2013a, Hartmann2014a}. There has been a significant increase in the area of subaerially exposed land within the region since the mid-Miocene associated with ongoing arc-continent collision between Australia and the Sunda-Banda arc system \cite{Molnar2015a, Hall2017a, Macdonald2019a}. Concurrently, after the Miocene Climatic Optimum, a cooling trend began ca. 15~Ma and accelerated over the past 4 million years (m.y.) leading to the development of Northern Hemisphere ice sheets \cite{Shackleton1984a, Zachos2001a}. Many hypotheses have been proposed to explain this cooling trend including changes in ocean/atmosphere circulation \cite{Haug1998a, Shevenell2004a, Molnar2015a}, a decrease in volcanic degassing \cite{Berner1983a}, or uplift in the Himalaya \cite{Raymo1988a, Galy2007a}. Here we seek to evaluate the hypothesis that emergence of the SEAI was a significant factor in driving long-term climatic cooling over the Neogene.

Over geologic time-scales, \COtwo enters Earth's ocean--atmosphere system primarily via volcanism and metamorphic degassing, and leaves primarily through the chemical weathering of silicate rocks and through organic carbon burial \cite{Kump2000a}. Chemical weathering delivers alkalinity and cations to the ocean which drives carbon sequestration through carbonate precipitation. Steady-state \pCOtwo is set at the \pCOtwo level at which \COtwo sinks are equal to sources. As \COtwo sinks increase and \pCOtwo falls, temperature decreases and the hydrological cycle is weakened, causing the efficiency of the silicate weathering sink to decrease until a new steady-state is achieved at lower \pCOtwo \cite{Kump1997a}.

Topography, climate, and lithology all effect chemical weathering. High-relief regions generally lack extensive regolith development, and thus tend to have reaction-limited weathering regimes that are more prone to adjust when climate changes \cite{Gabet2009a, West2012a, Maher2014a}. High physical erosion rates contribute to high chemical weathering fluxes in these high-relief regions \citep{Godderis2017b}. In warm and wet regions, mineral dissolution kinetics are faster leading to enhanced chemical weathering \cite{Lasaga1994a, West2012a}. Mafic rocks have higher Ca and Mg concentrations and dissolution rates than felsic rocks, and thus have the potential to more efficiently sequester carbon through silicate weathering \cite{Dessert2003a}. These factors have led to the proposal that arc-continent collisions, which create steep landscapes of mafic lithologies, within the tropical rain belt have been important in enhancing global weatherability, lowering atmospheric \pCOtwo, and initiating glacial climate over the past 520~m.y. \cite{Jagoutz2016a, Swanson-Hysell2017a, Macdonald2019a} and perhaps in the Neoproterozoic as well \cite{Park2020a}.

Quantitatively estimating the magnitude of decrease in steady-state \pCOtwo associated with the emergence of a region with a high carbon sequestration potential, such as the SEAI, requires constraints on changing tectonic context and accounting of associated feedbacks. As this region emerges, the total global silicate weathering flux will transiently exceed the volcanic degassing flux, causing \pCOtwo to initially decline until a new steady-state is established where the total magnitude of the \COtwo sinks is the same as before the change. However, the sensitivity of the silicate weathering flux in any particular location to this change in \pCOtwo is variable and dependent on the specific topography, climate, and lithology at that location. Furthermore, how regional climate responds to this change in \pCOtwo is itself spatially variable. Therefore, the magnitude of \pCOtwo change that is required to balance the total global silicate weathering flux with the volcanic degassing flux will depend on the specific spatial distribution of topography, climate, and lithology at the time of emergence. As a result, any attempt to meaningfully estimate the decrease in steady-state \pCOtwo associated with emergence of the SEAI must model spatially resolved climatology and silicate weathering fluxes in tandem and account for the spatial distribution of the factors that affect these inter-connected systems.

@@

To estimate the decrease in steady-state \pCOtwo associated with the increase of subaerially exposed land area in the SEAI, we use the global spatially resolved GEOCLIM model \cite{Godderis2017c}. GEOCLIM estimates changes in steady-state \pCOtwo associated with coupled changes in erosion, chemical weathering, and climatology by linking a silicate weathering model to climate model runs at multiple \pCOtwo levels.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{Figures/regolith_schematic.pdf}
    \caption{A schematic representation of the silicate weathering component of GEOCLIM in a single profile at steady-state. A rock particle leaves the unweathered bedrock with production rate $P_{r}$, and transits vertically through a regolith of height $h$. Regolith production and physical erosion ($E_{p}$) are equal at steady-state. As a particle transits upwards, some fraction of the primary phases ($x$) are chemically weathered ($W$), with the flux of dissolved Ca+Mg being $W$ multiplied by the concentration of Ca+Mg in unweathered bedrock ($\chi_{CaMg}$). Details of the formulation for the silicate weathering component of GEOCLIM can be found in \MM.}
    \label{fig:regolith_schematic}
\end{figure}

The silicate weathering component of GEOCLIM calculates \COtwo consumption resulting from silicate weathering for subaerially exposed land. We assume that Ca and Mg are the only cations that consume \COtwo over geologic time-scales, such that each mole of Ca or Mg that is dissolved by silicate weathering consumes one mole of \COtwo. While reverse weathering is another potential sink for Ca or Mg \citep{Michalopoulos1995a}, its parameterization is unclear and it has been interpreted to be a relatively minor flux in the Cenozoic \citep{Isson2018a}, and so we do not include it in our model. In previous versions of the model, silicate weathering was a function of temperature and runoff only, and all bedrock was assigned identical chemical compositions \cite{Godderis2017c}. More recent versions of GEOCLIM implement regolith development and soil shielding (Fig. \ref{fig:regolith_schematic}), which introduces a dependence on erosion rate (and therefore topographic slope) \cite{Maffre2018a}. While this introduction of regolith development into GEOCLIM is important for assessing the impact of tropical arc-continent collisions on \pCOtwo, the relatively high Ca+Mg concentration in arc rocks relative to other lithologies must also be considered.

@@

The values of four parameters within the silicate weathering component that modify the dependence of silicate weathering on temperature, runoff, erosion, and regolith thickness are poorly constrained. Rather than prescribing values, we allow these parameters and the Ca+Mg concentration of metamorphic and siliciclastic lithologies to vary within reasonable bounds (\SI; Table S2). This approach results in 93,600 unique combinations of these six parameters. For each combination, we compute spatially resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available \cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating parameter combinations that result in low $r^{2}$, 573 parameter combinations remain (\SI; Fig. S2). The resulting global \COtwo consumption of these filtered model runs all overlap with independently derived estimates of the global \COtwo degassing flux \cite{Gerlach2011a}, as they should given that the long-term carbon cycle is in steady state (\SI; Fig. S2).

@@

Having calibrated the silicate weathering component of GEOCLIM, we use it to estimate the decrease in steady-state \pCOtwo associated with emergence of the SEAI. For the climate model component, we use temperature and runoff from a subset of the GFDL CM2.0 experiments \cite{Delworth2006b} (\SI). These experiments are well-suited for this analysis because all non-\COtwo forcings are held constant at values representative of pre-industrial conditions, allowing the effect of changing \pCOtwo on climatology to be isolated. Furthermore, the experiments were run long enough for the final system to approximate steady-state.

@@

\begin{figure*}[h]
    \centering
    \includegraphics[width=1\linewidth]{Figures/scenario_pCO2.pdf}
    \caption{Steady-state \pCOtwo estimates from GEOCLIM for the various scenarios discussed in the text. For each of the seven scenarios, each point represents an estimate from one of the 573 unique parameter combinations that most closely matched estimates of present-day \COtwo consumption in 80 watersheds around the world (\SI). The box encloses the middle 50\% of the \pCOtwo estimates (i.e. the interquartile range), and the notch represents the median with its 95\% confidence interval. The whiskers extend to the 2.5 and 97.5 percentile values. Glaciation thresholds \cite{DeConto2008a} are shown on the x-axis.}
    \label{fig:scenario_pCO2}
\end{figure*}

To determine the position of paleoshorelines in the SEAI over the past 15~m.y., we use terrestrial and marine sedimentary deposits (Fig. \ref{fig:shoreline_growth}; \SI). The paleoshoreline data indicate that the Sunda-Banda Arc and New Guinea are primarily responsible for the increase in area since 15~Ma. Exhumation of the modern Sunda-Banda Arc is the result of ongoing arc-continent collision with the Australian Plate \cite{Harris2006a}. Most of Sumatra and Java along with the non-volcanic islands of the Outer Banda Arc were elevated above sea level after 5~Ma \cite{Hall2013b}. In New Guinea, emergence in the mid-Miocene is associated with collision between the Melanesian Arc and Australia's distal margin \cite{Cloos2005a}, which drove exhumation of the Irian-Marum-April Ophiolite Belt. Exhumation accelerated over the past 4~m.y. in the New Guinea Central Range due to slab-breakoff and buoyant uplift, and in eastern New Guinea due to jamming of the north-dipping subduction zone \cite{Cloos2005a}. We also include changes in areas of presently submerged continental shelves such as the Sunda Shelf that were previously exposed (\SI; Fig. S5). These tectonic drivers and others throughout the region led to progressive emergence over the past 15~m.y. that accelerated following 5~Ma (Fig. \ref{fig:shoreline_growth}B). This trend mirrors broad cooling over the Neogene that resulted in the initiation of Northern Hemisphere ice sheets (Fig. \ref{fig:shoreline_growth}C).

We use GEOCLIM to estimate \pCOtwo associated with the reconstructed subaerial extent of the SEAI at ca. 15, 10, and 5~Ma (``paleo-SEAI'' scenarios; Fig. \ref{fig:scenario_pCO2}). Because we use a climate model forced with modern geography, the position of the tectonic blocks remain fixed. Although there has been motion of these tectonic blocks since 15~Ma, they have remained within tropical latitudes such that this fixed scenario is a good approximation of the paleogeography (\SI; Fig. S8). We also test an end-member scenario, in which all islands associated with arc-continent collision in the region are removed (``removed SEAI'' scenario; Fig. \ref{fig:scenario_pCO2}).

@@

Using the 573 unique parameter combinations, the ``paleo-SEAI'' scenarios resulted in 526--678~ppm for 15~Ma, 457--516~ppm for 10~Ma, and 391--434~ppm \pCOtwo for 5~Ma (Fig. \ref{fig:scenario_pCO2}). These results indicate a progressive decrease in \pCOtwo over the Neogene associated with the emergence of the SEAI, and suggest that without this emergence, pre-industrial \pCOtwo would have been $\sim$526--678~ppm. We note that these ``paleo-SEAI'' scenarios do not account for Neogene changes outside of the SEAI (e.g. changes in ocean/atmosphere circulation, volcanic degassing, and weathering fluxes elsewhere on Earth, discussed in \textit{Alternative Mechanisms for Neogene Cooling}). Therefore, these results do not imply that \pCOtwo would have been observed to be 526--678~ppm at 15~Ma, but rather that emergence of the SEAI alone is responsible for decreasing \pCOtwo by 240--392~ppm (526--678~ppm minus 286~ppm) over the past 15~m.y..

Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation (\SI; Fig. S9). The \pCOtwo values from the 5~Ma ``paleo-SEAI'' scenario overlap with many proxy-based estimates \cite{Foster2017a} as well as values that emerge from approaches that assimilate climate and ice sheet model output with benthic \dO data \cite{van-de-Wal2011a, Berends2020a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early to mid-Miocene, indicating that the increase in subaerially exposed land area and tectonic topography of the SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo associated with emergence of the SEAI are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

The results of our ``paleo-SEAI'' scenarios highlight the importance of the combination of topography, runoff, and lithology in setting Earth's climate state. To independently explore the effect of the modern-day surface exposure of low-relief basaltic lavas on steady-state \pCOtwo \cite{Kent2013a}, we replace mafic volcanics associated with the Deccan Traps, Ethiopian Traps, and Columbia River Basalts with the Ca+Mg concentration of bulk continental crust in GEOCLIM (Fig. \ref{fig:scenario_pCO2}). The resulting \pCOtwo is $<$500~ppm, indicating that, while the presence of mafic rocks in these igneous provinces modulates steady-state \pCOtwo as has been suggested to be important for Paleogene cooling \cite{Kent2013a}, its contribution is less than that of the SEAI, which have high-relief and a wet tropical climate in conjunction with high Ca+Mg lithologies.

Previous work has estimated that the decrease in \pCOtwo since 5~Ma associated with the emergence of the SEAI and enhanced silicate weathering is $\sim$19~ppm \cite{Molnar2015a}, in which case their emergence would be a relatively minor contributor to Neogene cooling. This 19~ppm estimate was obtained using a 0-D equation that assumes a direct linear relationship between mean global temperature and changes in weathering-rate-weighted land area, scaled by a factor that is intended to account for the influence of both runoff and temperature. \pCOtwo was then estimated from the calculated temperature using a simple energy balance equation. However, the relationship between mean global temperature (or \pCOtwo) and weathering-rate-weighted land area is not linear. Furthermore, this simple linear relationship ignores spatial variability in topography and climatology, and only crudely accounts for spatial variability in lithology. In fact, the 19~ppm estimate is closer in magnitude to the decrease in \pCOtwo that we estimate if mafic volcanics associated with the Deccan Traps (a relatively flat area outside of the warm and wet tropics) are replaced with the Ca+Mg concentration of bulk continental crust (22--70 ppm; Fig. \ref{fig:scenario_pCO2}). The significant difference in steady-state \pCOtwo estimated between the ``removed Deccan Traps'' scenario and the ``paleo-SEAI'' scenarios (Fig. \ref{fig:scenario_pCO2}) demonstrates that considering changes in the spatial distribution of lithologies alone is not adequate for estimating changes in steady-state \pCOtwo. Instead, spatially varying topography and climatology significantly modulates silicate weathering rates, and must be accounted for when estimating \pCOtwo change associated with paleogeographic change.

@@

\section*{The Geologic Carbon Cycle}
